Parsing Postfix log files

### 2007-04-15 doc/logparser.tex
Tweaks suggested by chktex.

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator; every Real-time Black List returns a
different explanatory message; policy servers may have different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database.  The gathered data can then be used to
optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying high volume uses (e.g.  customer newsletters) and restricting
those uses to off-peak times; as a base for billing customers on a shared
server; etc.

Section~\ref{background} provides background information about: the
motivation for this project; the Postfix Mail Transport Agent; a short
overview of the parser; the assumptions inherent in the design of the
algorithm and program; using a database schema as an API; and a comparison
against other Postfix log parsers.

@@

    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
        over-view of postfix activity, with just enough detail to give the
        administrator a "heads up" for potential trouble spots.}
        \url{http://jimsun.linxnet.com/postfix_contrib.html}

@@

    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
        in C. It designed to scan qmail, postfix, sendmail and exim logfile
        and produce usage statistics in HTML format for viewing through a
        browser. It produces Top domains output according to Sender,
        Receiver, Total mails and bytes; it keeps your main domain mail
        statistics with regard to Days Top Domain, Top Users values for per
        day, per month and years.}  \newline
        \url{http://www.enderunix.org/isoqlog/}

    \item [AWStats] \textit{AWStats is a free powerful and featureful tool
        that generates advanced web, streaming, ftp or mail server
        statistics, graphically.}

@@

    \item [Log analyser - throughput monitor] This utility tracks the
        number of events which occurred over a particular time and warns if
        the frequency of events passes a certain threshold.  It's designed
        to provide real time alerts when dictionary attacks, mail loops or
        similar problems occur.  \newline
        \url{http://home.uninet.ee/~ragnar/throughput_monitor/}

    \item [Anteater] \textit{The Anteater project is a Mail Traffic
        Analyser. Anteater supports currently the logformat produced by
        Sendmail and by Postfix. The tool is written in 100\% C++ and is
        very easy to customize. Input, output, and the analysis are modular
        class objects with a clear interface. There are eight useful
        analyse modules, writing the result in plain ASCII or HTML, to
        stdout or to files.}

@@

    \item [Yet Another Advanced Logfile Analyser] uses a plugin based
        system to analyse log files and produce output reports.  The core
        code is merely 91 lines long, as all the parsing and report
        generation is handled by modules.  Using YAALA as a base would be
        only slightly less work as both input and output modules would need
        to be written; it may even be more work to implement the parser
        within the constraints of YAALA.

@@

    \item [Logparser/Lire] Lire is another general purpose log parser, but
        is developed under the GNU GPL.  Like other general purpose log
        parsers it doesn't extract enough data from Postfix logs to be
        worthwhile.  Development also seems to have stalled, with only a
        minor update since October 2004.  \newline
        \url{http://logreport.org/lire/}

    \item [Logrep] \textit{Logrep is a secure multi-platform framework for
        the collection, extraction, and presentation of information from
        various log files.}

>>>>

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises as a result of the facility to have rejection messages
defined by the administrator; every Real-time Black List returns a
different explanatory message; policy servers may have different messages
depending on the characteristics of the connection; there are many ways in
which the log lines may differ.  This paper documents the difficult process
of parsing Postfix logs, presenting a program which parses logs and places
the resulting data into a database.  The gathered data can then be used to
optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying high volume uses (e.g.\ customer newsletters) and restricting
those uses to off-peak times; as a base for billing customers on a shared
server; etc.

Section~\ref{background} provides background information about: the
motivation for this project; the Postfix Mail Transport Agent; a short
overview of the parser; the assumptions inherent in the design of the
algorithm and program; using a database schema as an API\@; and a
comparison against other Postfix log parsers.

@@

    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
        over-view of postfix activity, with just enough detail to give the
        administrator a ``heads up'' for potential trouble spots.\/}
        \url{http://jimsun.linxnet.com/postfix_contrib.html}

@@

    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
        in C. It designed to scan qmail, postfix, sendmail and exim logfile
        and produce usage statistics in HTML format for viewing through a
        browser. It produces Top domains output according to Sender,
        Receiver, Total mails and bytes; it keeps your main domain mail
        statistics with regard to Days Top Domain, Top Users values for per
        day, per month and years.\/}  \newline
        \url{http://www.enderunix.org/isoqlog/}

    \item [AWStats] \textit{AWStats is a free powerful and featureful tool
        that generates advanced web, streaming, ftp or mail server
        statistics, graphically.\/}

@@

    \item [Log analyser --- throughput monitor] This utility tracks the
        number of events which occurred over a particular time and warns if
        the frequency of events passes a certain threshold.  It's designed
        to provide real time alerts when dictionary attacks, mail loops or
        similar problems occur.  \newline
        \url{http://home.uninet.ee/~ragnar/throughput_monitor/}

    \item [Anteater] \textit{The Anteater project is a Mail Traffic
        Analyser. Anteater supports currently the logformat produced by
        Sendmail and by Postfix. The tool is written in 100\% C++ and is
        very easy to customize. Input, output, and the analysis are modular
        class objects with a clear interface. There are eight useful
        analyse modules, writing the result in plain ASCII or HTML, to
        stdout or to files.\/}

@@

    \item [Yet Another Advanced Logfile Analyser] uses a plugin based
        system to analyse log files and produce output reports.  The core
        code is merely 91 lines long, as all the parsing and report
        generation is handled by modules.  Using YAALA as a base would be
        only slightly less work as both input and output modules would need
        to be written; it may even be more work to implement the parser
        within the constraints of YAALA\@.

@@

    \item [Logparser/Lire] Lire is another general purpose log parser, but
        is developed under the GNU GPL\@.  Like other general purpose log
        parsers it doesn't extract enough data from Postfix logs to be
        worthwhile.  Development also seems to have stalled, with only a
        minor update since October 2004.  \newline
        \url{http://logreport.org/lire/}

    \item [Logrep] \textit{Logrep is a secure multi-platform framework for
        the collection, extraction, and presentation of information from
        various log files.\/}
